A hierarchical approach for modelling X-ray beamlines. Application to a coherent beamline.

### 2019-05-07 paper-hierarchical.tex
Updates from Overleaf

The paper is structured as follows: we first introduce in Section~\ref{Beamline description} the main parameters of the beamline and explain the particular choice of the parameters used in the calculations. Section~\ref{Simulations} is the central part of the paper, and describes the different methods used to analyse the beamline with increasing complexity, setting a hierarchy of methodologies that can be exported to the simulations of any beamline. In Section \ref{level0} we calculate with simple methods approximate values of focal size, flux and coherent fraction. Accuracy can improved using geometrical methods in Section \ref{level1} that allow a more complete simulation of the beamline optics within certain limits which are related, in this example, to high demagnification and high coherence of the studied beamline. The wave optics section~\ref{level2} includes different models of increasing complexity as the required accuracy of the results becomes more demanding. The simulation methods are presented in a hierarchical approach:  they provide more and better results but the price to pay is to deal with more complex and computationally expensive tools.

@@

ID16A \cite{ID16A} is a beamline that provides a nanometric, highly coherent beam with broad energy-bandpass. It is dedicated to nanofocusing applications. \todo{please summarize some use and application}. The beamline can operate at three fixed photon energies: 11200 eV, 17225 eV and 33600 eV. The straight section in the magnet lattice houses three undulators: an in-vacuum undulator of $\lambda_u=26$~mm period, and two other conventional undulators of 18 and 22.4 mm period. The undulator parameters and $K$-values at the resonances for the energies in use are shown in Table \ref{tab:Undulators}.

@@

\resizebox{\textwidth}{!}{%
\begin{tabular}{lcccccc}
name  & $\lambda_u$ [mm] & $L$ [m] & N & $K$ (11200 eV) & $K$ (17225 eV) & $K$ (33600 eV) \\ 
\input{tableUndulator.txt}
\end{tabular}%
}
\end{table}

@@

For all simulations presented hereafter, we selected the configuration using a single undulator U18.3 placed in the center of the straight section and tuned to its first harmonic at a photon energy of $E=17225$ eV (Table \ref{tab:Undulators}). This section is divided in three subsections, that present three main levels of simulations: simple calculations that can be made by hand (section \ref{level0}), ray-tracing simulations (section \ref{level1}) and wave optics simulations (section \ref{level2}).

@@

The undulator natural photon source size and divergence are given by \cite{elleaume}:
\begin{align}
    \label{eq:photon small sigmas}
    \sigma_u=\frac{2.74}{4\pi}\sqrt{\lambda L}\approx \sqrt{\frac{\lambda L}{2 \pi^2}}  && \sigma_u' = 0.69\sqrt{\frac{\lambda}{L}}\approx \sqrt{\frac{\lambda}{2 L}},
\end{align}
where $\lambda$ is the emitted wavelength and $L=N\Lambda_u$ is the undulator length, with $\Lambda_u$ as the magnetic period of the undulator and $N$, the number of magnetic periods. The electron beam usually presents different horizontal and vertical beam emittances, but the undulator natural source size or divergence presents radial symmetry at the central cone (excluding polarization). The photon source size and divergence are approximately given by the convolution of the electron beam sizes and the undulator natural sizes. If we consider Gaussian distributions, the photon source is
\begin{align}
\label{eq:photon big sigmas}
\Sigma_{h,v}=\sqrt{\sigma_{e(h,v)}^2 + \sigma_u^2} && \Sigma_{h,v}'=\sqrt{\sigma_{e(h,v)}'^2 + \sigma_u'^2}.
\end{align}

@@

The HYBRID method \cite{hybrid} uses geometric ray tracing combined with wavefront propagation. The code computes diffraction effects when the beam is clipped by an aperture or by the finite size of the optics and can also simulate the effect of mirror figure errors when diffraction is present. The method can be applied to an entire beamline by simulating each optical element iteratively under the hypothesis of the validity of far field propagation in all intermediate propagations. If this is not the case, a near field version is available for investigating imaging by individual optics. The interest of the hybrid method is to assess in a fast way the influence of beam coherence because the code is considerably faster than the the full methods for dealing with the partial coherence of the synchrotron beam that are described below. Fig.~\ref{fig:ray tracing}b presents the results using the hybrid model for the beamline with ideal elliptical optical surfaces. When compared with the pure ray tracing in Fig.~\ref{fig:ray tracing}a it can be appreciated that the spot is broadened. This is more evident in the vertical direction, because the smaller focal size. The spot size is about 7-8 nm in V for both storage ring lattices, and in horizontal is about 15$\mu$m for ESRF-High$\beta$ and 11 nm for EBS. It is interesting to note the almost round spot that will be obtained with the new EBS lattice.

The surface finish of the optical elements is another limiting effect for determining the size and intensity of the beam at the image position. It is possible to infere the effect of mirror imperfections and in particular slope errors by only knowing some statistical parameters such as the slope error RMS and then building synthetic surface profiles to be used in the simulations. This is the typical mode to operate when designing a new beamline where the mirrors do not exist yet. However, in our case the mirrors already exist and have been extensively characterize at the ESRF metrology laboratory. It is therefore highly recommended to use metrology data from real mirrors, when possible. When not available, one can use data from existing mirrors from a database (e.g., \cite{dabam}) and extrapolate some of the effects to the hypothetical new mirrors. The metrology data of the mirrors for the beamline under study are presented in Appendix~\ref{appendix:metrology}. These experimental data are used in all the simulations in this paper. Simulations using the hybrid model including slopes errors are shown in Fig.~\ref{fig:ray tracing}c. It can be observed a small broadening of the peak due to slope errors, but, more importantly, a new satellite structure close to the vertical intensity peak is produced by the slope errors.

@@

In this section the coherence properties of the beamline are analysed using, again, different methodologies classified by complexity and computer resources. We first analyze the beamline supposing a fully coherent beam. Subection~\ref{wofry} shows a very simpified 1D model. Subsection~\ref{srw_se} used a full 2D model with accurate simulation of the undulator emission but without considering storage ring emittance. These simplifications permit to address diffraction effects by apertures and slope errors. Then we introduce methods to treat the partial coherence originated by the storage ring emittance. Two variants are shown, a Monte Carlo approach (subsection~\ref{srw_me}) that permits us to calculate in great detail the intensity dostribution of the focused beam, and a decomposition of the radiation in coherent modes (subsection~\ref{comsyl}) that allows to evaluate quantitatively the quality of the coherent beam.

@@

The very long distance source-sample combined with a small source (in particular for EBS) makes this beamline suitable for experimental techniques requiring a highly coherent beam. Indeed, it has been shown that the clipping of the beam, mainly due to the acceptance of the KB mirrors produces a significant broadening of the focal spot.  As discussed in Section~\ref{level0}, very simple calculation of the diffraction limit can be sone using a convergent spherical beam clipped by a slit of 400$\times$400$ \mu$m$^2$ placed at 5cm upstream the focus (at the second KB mirror position). The slit vertical size corresponds roughly to the horizontal acceptance of the seconf KB mirror, and it is approximately the same in vertical. A simple simulation of diffraction by a square slit using OASYS is shown in Fig.~\ref{wofry400um}. It reveals a spot of less than 10 nm, thus confirming the analytical predictions, and qualifying this beamline to be a diffraction limited beamline.

@@

\section{Summary and conclusions}

>>>>

The paper is structured as follows: we first introduce in Section~\ref{Beamline description} the main parameters of the beamline and explain the particular choice of the parameters used in the calculations. Section~\ref{Simulations} is the central part of the paper, and describes the different methods used to analyse the beamline with increasing complexity, setting a hierarchy of methodologies that can be exported to the simulations of any beamline. In Section \ref{level0} we calculate with simple methods approximate values of focal size, flux and coherent fraction. Accuracy can improved using geometrical methods in Section \ref{level1} that allow a more complete simulation of the beamline optics within certain limits which are related, in this example, to high demagnification and high coherence of the studied beamline. The wave optics section~\ref{level2} includes different models of increasing complexity as the required accuracy of the results becomes more demanding. \inblue{In the Discussion (Section \ref{discussion}) the results of the different simulation methods are commented. The hierarchical approach allows more accurate and better results but the price to pay is to deal with more complex and computationally expensive tools. Finally Section \ref{summary} summarizes the results.}

@@

ID16A \cite{ID16A} is a beamline that provides a nanometric, highly coherent beam with broad energy-bandpass. It is dedicated to nanofocusing applications. \todo{please summarize some use and application}. The beamline can operate at two fixed photon energies: 17050 eV and 33600 eV. The straight section in the magnet lattice houses four undulators, two 18 mm and another two of 22.4 mm period. The undulator parameters and $K$-values at the resonances for the energies in use are shown in Table \ref{tab:Undulators}.

@@

\resizebox{\textwidth}{!}{%
\begin{tabular}{lcccccc}
name  & $\lambda_u$ [mm] & $L$ [m] & N & $K$ (11200 eV) & $K$ (17050 eV) & $K$ (33600 eV) \\ 
\input{tableUndulator.txt}
\end{tabular}%
}
\end{table}

@@

For all simulations presented hereafter, we selected the configuration using a single undulator U18.3 placed in the center of the straight section and tuned to its first harmonic at a photon energy of $E=17225$ eV (\inblue{this value is not identical but close to the usual operation value shown in }Table \ref{tab:Undulators}). This section is divided in three subsections, that present three main levels of simulations: simple calculations that can be made by hand (section \ref{level0}), ray-tracing simulations (section \ref{level1}) and wave optics simulations (section \ref{level2}).

@@

The undulator natural photon source size and divergence are given by \cite{elleaume}:
\begin{align}
    \label{eq:photon small sigmas}
    \sigma_u=\frac{2.74}{4\pi}\sqrt{\lambda L}\approx \frac{1}{2  \pi}\sqrt{2 \lambda L}  && \sigma_u' = 0.69\sqrt{\frac{\lambda}{L}}\approx \sqrt{\frac{\lambda}{2 L}},
\end{align}
where $\lambda$ is the emitted wavelength and $L=N\Lambda_u$ is the undulator length, with $\Lambda_u$ as the magnetic period of the undulator and $N$, the number of magnetic periods. The electron beam usually presents different horizontal and vertical beam emittances, but the undulator natural source size or divergence presents radial symmetry at the central cone (excluding polarization). The photon source size and divergence are approximately given by the convolution of the electron beam sizes and the undulator natural sizes. If we consider Gaussian distributions, the photon source is
\begin{align}
\label{eq:photon big sigmas}
\Sigma_{h,v}=\sqrt{\sigma_{e(h,v)}^2 + \sigma_u^2} && \Sigma_{h,v}'=\sqrt{\sigma_{e(h,v)}'^2 + \sigma_u'^2}.
\end{align}

@@

The HYBRID method \cite{hybrid} uses geometric ray tracing combined with wavefront propagation. The code computes diffraction effects when the beam is apertured and can also simulate the effect of mirror figure errors when diffraction is present. The method can be applied to an entire beamline by simulating each optical element iteratively under the hypothesis of the validity of far field propagation in all intermediate propagations. If this is not the case, a near field version is available for investigating imaging by individual optics. The interest of the hybrid method is to assess in a fast way the influence of beam coherence because the code is considerably faster than the the full methods for dealing with the partial coherence of the synchrotron beam that are described below. Fig.~\ref{fig:ray tracing}b presents the results using the hybrid model for the beamline with ideal elliptical optical surfaces. When compared with the pure ray tracing in Fig.~\ref{fig:ray tracing}a it can be appreciated that the spot is broadened. This is more evident in the vertical direction, because of the smaller focal size. The spot size is about 7-8 nm in V for both storage ring lattices, and in horizontal is about 15$\mu$m for ESRF-High$\beta$ and 11 nm for EBS. It is interesting to note the almost round spot that will be obtained with the new EBS lattice.

The surface figure of the optical elements is another limiting effect for determining the size and intensity of the beam at the image position. It is possible to infer the effect of mirror imperfections and in particular slope errors by only knowing some statistical parameters such as the RMS slope error  and then building synthetic surface profiles to be used in the simulations. This is the typical approach used when designing a new beamline where the mirrors do not exist yet. However, in our case the mirrors already exist and have been extensively characterized at the ESRF optical metrology laboratory. It is therefore highly recommended to use metrology data from real mirrors, when possible. When not available, one can use data from existing mirrors from a database (e.g., \cite{dabam}) and extrapolate some of the effects to the hypothetical new mirrors. The metrology data of the mirrors for the beamline under study are presented in Appendix~\ref{appendix:metrology}. These experimental data are used in all the simulations in this paper. Simulations using the hybrid model including slope errors are shown in Fig.~\ref{fig:ray tracing}c. A small broadening of the peak due to slope errors can be observed, but, more importantly, a new satellite structure close to the vertical intensity peak is produced by the slope errors.

@@

In this section the coherence properties of the beamline are analysed using, again, different approaches  distinguished by their complexity and necessary computer resources. We first analyze the beamline supposing a fully coherent beam. Subection~\ref{wofry} shows a very simplified 1D model. Subsection~\ref{srw_se} used a full 2D model with accurate simulation of the undulator emission but without considering storage ring emittance (i.e. again using a fully coherent beam). These simplifications permit the simulation of diffraction effects by apertures and slope/figure errors. Subsequently we introduce methods to treat the partial coherence originated by the storage ring emittance. Two variants are shown, a Monte Carlo approach (subsection~\ref{srw_me}) that permits us to calculate in great detail the intensity distribution of the focused beam, and a decomposition of the radiation in coherent modes (subsection~\ref{comsyl}) that allows   quantitative evaluation of the coherent beam characteristics.

@@

The very long  source-{\color{green}\sout{sample} optics} distance combined with a small source (in particular for EBS) makes this beamline suitable for experimental techniques requiring a highly coherent beam. Indeed, it has been shown that the clipping of the beam, mainly due to the acceptance of the KB mirrors produces a significant broadening of the focal spot.  As discussed in Section~\ref{level0}, very simple calculation of the diffraction limit can be done using a convergent spherical beam clipped by a slit of 400$\times$400$ \mu$m$^2$ placed  5cm upstream the focus (at the second KB mirror position). The slit vertical size corresponds roughly to the horizontal acceptance of the seconf KB mirror, and it is approximately the same in vertical. A simple simulation of diffraction by a square slit using OASYS is shown in Fig.~\ref{wofry400um}. It reveals a spot of less than 10 nm, thus confirming the analytical predictions, and qualifying this beamline to be a diffraction limited beamline.

@@

\section{Summary and conclusions}
\label{summary}
